Robotic Bicycle Balance Assistance Reduces Probability of Falling When Subjected to Mechanical Perturbations

### 2024-03-22 main.tex
Added more in measurements and control algorithm sections, added references.

\usepackage{siunitx}

\title{Robotic Bicycle Balance Assist Reduces Probability of Falling When
Subjected to Mechanical Perturbations}

@@

\subsection{Balance Assist Control}
We use a forward speed gain scheduled proportional roll rate positive feedback
controller to stabilize the bicycle. Scaling the proprtional feedback gain
linearly with respect to speed stabilizes the normally unstable weave mode of
the bicycle from speeds of X~\si{\kilo\meter\per\hour} to
X~\si{\kilo\meter\per\hour} as shown in Figure~\ref{fig:eigenvalues}.
%
\begin{figure}
  \label{fig:eigenvalues}
\end{figure}

\subsection{Perturbations}
%
We apply longitudinal forces to the ends of each handlebar using an adapted
Bump'Em system~\cite{todo} making use of four motors working in tandem. The
four motors are programmed to apply a light force at all times to keep the
ropes taught and to track a commanded force profile using a PID controller
running on a microprocessor (Arduino XXX, Italy). We measure the force applied
by each motor at the handlebar via four inline load cells with a max load of
XX~\si{\newton}. The commanded force profiles are designed to apply an external
pulsive torque to the front assembly (handlebars, forks, wheel) at magnitudes
varying from X to X. The four motors are arranged at the four corners of a 1
meter wide treadmill (make, Company, City) that can reach speeds of
18~\si{\kilo\meter\per\hour}. The general design is described in detail in the
Van Velde's MSc thesis~\cite{vanvelde2022}. Our modifications to that described
in the thesis include simplifying the controller with a inexpensive
microcontoller and the use of a simple safety harness.

@@

\subsection{Subjects}
%
We recruited X healthy young adults to participate in the experiments. X
subjects performed the experiments at 6~\si{\kilo\meter\per\hour} and X
subjects performed the experiments at 10~\si{\kilo\meter\per\hour}. The
10~\si{\kilo\meter\per\hour} experiments occured X weeks before the
6~\si{\kilo\meter\per\hour}.

\subsection{Measurements}

@@

>>>>

\usepackage{amsmath}
\usepackage[margin=25mm]{geometry}
\usepackage{siunitx}  % use for all units
\usepackage{todonotes}

\title{Robotic Bicycle Balance Assistance Reduces Probability of Falling When
Subjected to Mechanical Perturbations}

\author{Marten T. Haitjema \and Leila Alizadehsaravi \and Jason K. Moore}

@@

\subsection{Balance Assist Control}
We use a forward speed \(v\) gain scheduled proportional roll rate positive
feedback controller to stabilize the bicycle. The commanded steer torque
\(T_\delta\) from the steer motor follows
\begin{align}
  T_\delta = -k(4.7\si{\meter\per\second} - v)\dot{\phi}
\end{align}
where 4.7~\si{\meter\per\second} is the weave critical speed predicted from the
open loop bicycle rigid rider dynamics. We use \(k=8\) as the low gain value
and \(k=10\) as a high gain value. Scaling the proprtional feedback gain
linearly with respect to speed stabilizes the normally unstable weave mode of
the bicycle from speeds of X~\si{\kilo\meter\per\hour} to
X~\si{\kilo\meter\per\hour} as shown in Figure~\ref{fig:eigenvalues}.

\todo[inline]{Show open loop bicycle rigid rider eigenvalue vs speed and then
the closed loop versions with the 8 and 10 gain settings. Three plots.}

\begin{figure}
  \label{fig:eigenvalues}
\end{figure}

\subsection{Perturbations}
%
We apply longitudinal forces to the ends of each handlebar using an adapted
Bump'Em system~\cite{Tan2020} making use of four motors working in tandem. The
four motors are programmed to apply a light force at all times to keep the
ropes taught and to track a commanded force profile using a PID controller
running on a microprocessor (Arduino XXX, Italy). We measure the force applied
by each motor at the handlebar via four inline load cells with a max load of
XX~\si{\newton}. The commanded force profiles are designed to apply an external
pulsive torque to the front assembly (handlebars, forks, wheel) at magnitudes
varying from X to X. The four motors are arranged at the four corners of a 1
meter wide treadmill (make, Company, City) that can reach speeds of
18~\si{\kilo\meter\per\hour}. The general design is described in detail in the
Van De Velde's MSc thesis~\cite{vandeVelde2022}. Our modifications to that
described in the thesis include simplifying the controller with a inexpensive
microcontoller and the use of a simple safety harness.

@@

\subsection{Subjects}
%
We recruited 26 able-bodied young adults (20-36 years old) to participate in
the experiments. The subjects all were confident in their cycling skills and
had cycled at least once in the last month. Eleven subjects performed the
experiments at 6~\si{\kilo\meter\per\hour} and fifteen subjects performed the
experiments at 10~\si{\kilo\meter\per\hour}. All subjects consented to the
experiment and could decline to continue at any time.  The study was approved
by Delft University of Technology's Human Research Ethics Council (#XXX).

\subsection{Two Experiments}
%
The 10~\si{\kilo\meter\per\hour} experiments occured X weeks before the
6~\si{\kilo\meter\per\hour}.

\subsection{Measurements}
%
We measure the time histories of the Bump'Em delivered perturbation forces and
the bicycle's steer angle, roll angle, roll angular rate, and forward speed.
%
\begin{table}
  \caption{Raw measurements}
  \begin{tabular}{llll}
    Measure & Variable & Units & Sensor \\
    Perturbation Order Number & \(j\) & integer & NA\\
    Perturbation Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear & \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & Load cell\\
    Steer Angle & \(\delta\) & \si{\degree} & encoder \\
    Roll Angle & \(\phi\) &  \si{\degree} & Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) &  \si{\degree\per\second} & rate gyroscope \\
    Balance Assist Gain & \(k\) & TODO & NA \\
    Bicycle Speed & \(v\) & Meters per second & wheel encoder \\
  \end{tabular}
\end{table}

Based on the findings in \cite{Haitjema2023}, we calculate several variables.
We use the angular impulse \(L\) due to the perturbation forces over a
0.3~\si{\second} duration to characterize the magnitude of delivered
pertrubation. The duration is selected based on the duration of the commanded
force.

\begin{align}
  F_r = & F_{rf} - F_{rr} \\
  F_l = & F_{lf} - F_{lr} \\
  L = & \int_{0\si{\second}}^{0.3\si{\second}} l/2(F_r + F_l) dt
\end{align}

We characterize the skill of the rider by calculating the standard deviation of
the steer angle \(\sigma_\delta\) over the 90~\si{\second} period they were
asked to maintain their position at the center of the treadmill. At the
intiation of each perturbation we log the steer and roll angles to characterize
the configuration of the bicycle when perturbed. The gain setting on the
balance assist controller indicates if the assistance is off \(k=0\) or on at
two different levels: low \(k=8\) or high \(k\=10\). A recovery from the
perturbation is successful if the person neither put their foot down onto the
treadmill surface or either wheel of the bicycle exits the width of the
treadmill belt. We record this as a binary variable \(f\).

\begin{table}
  \begin{tabular}{lll}
    \(L\) & \si{\newton\meter\second} & angular impulse of the applied
    perturbation torque \\
    \(j\) & integer & order number of perturbation \\
    \(\sigma_\delta\) & standard deviation of steer angle during Phase X \\
    \(k\) & TODO & gain of balance assist control \\
    \(\delta_0\) & \si{\degree} & steer angle at start of perturbation \\
    \(\phi_0\) & \si{\degree} & roll angle at start of perturbation \\
    \(v\) & \si{\meter\per\second} & forward speed (6 or 10) \\
    \(f\) & boolean & outcome of perturbation: did not fall, did fall \\
  \end{tabular}
\end{table}

\subsection{Statistics}

@@

\bibliographystyle{plain}
\bibliography{references}
